Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study

### 2024-03-20 manuscript.Rmd
20/3

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS. An explanation for the
increased odds of OFI in patients with increased RTS could be that the
patients with low RTS values - with low GCS, RR and SBP also are the
once that are in a bad state upon arrival --- upon which emergency
procedures e.g. intubation have been carried out rapidly due to the
patients being visibly ill from the beginning. The patients who are
being missed and where certain procedures take longer time before being
performed are the patients who seem stable upon arrival, but who later
become worse --- the patients with higher RTS scores. In Albaaj's study
respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
mmHg, and Glasgow Coma Scale 9-12 were significantly related to
increased odds of OFI in the unadjusted analysis, but not in the
adjusted analysis among patients from the Swedish Trauma Registry [@12].
All of these variables equal medium to higher RTS values --- capturing
the group of patients who's vital parameters are negatively affected by
the trauma, but who are relatively stable in this group of patients.
This creates a sense of false security, leading to increased time until
necessary emergency procedures are performed.

#Se om det finns källa som pekar på vilken grupp av patienter som bör få
mer fokus på IVA?

@@

#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
vad orsakerna är

@@

#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
bedömningen av OFI

The variable on duty was not significant in either analyses indicating
that the time of arrival of the patients did not affect their medical
treatment. Trying to reduce number of OFI can therefore not be solved by
increasing medical resources outside working hours.

#Lägga in artikel som pekar på om duty påverkar handläggning eller inte

@@

#Artikel om hur ISS bedöms kunna användas i samanhang av trauma

### Limitations

@@

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna. This leads to a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI later on in the hospitalization
process.

#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)

@@

@@

>>>>

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS. An explanation for the
increased odds of OFI in patients with increased RTS could be that the
patients with low RTS values - with low GCS, RR and SBP also are the
once that are in a bad state upon arrival --- upon which emergency
procedures e.g. intubation have been carried out rapidly due to the
patients being visibly ill from the beginning. The patients who are
being missed and where certain procedures take longer time before being
performed are the patients who seem stable upon arrival, but who later
become worse --- the patients with comparatively higher RTS scores. In
Albaaj's study respiratory rate \>29, systolic blood pressure 50-75 mmHg
and 76-89 mmHg, and Glasgow Coma Scale 9-12 were significantly related
to increased odds of OFI in the unadjusted analysis, but not in the
adjusted analysis among patients from the Swedish Trauma Registry [@12].
All of these variables equal medium to higher RTS values --- capturing
the group of patients who's vital parameters are negatively affected by
the trauma, but who are relatively stable in the group of trauma
patients. This creates a sense of false security, leading to increased
time until necessary emergency procedures are performed.

A previous study on pediatric patients in South Africa showed that RTS
was not predictive of mortality between high-risk and low-risk patients
[@21]. This could indicate... .

@@

Previous studies have shown that prolonged ICU stay is a risk factor for
increased mortality and morbidity in trauma patients [@22, @26].
Something to consider is the reason for the lengthened stay in the ICU -
studies have shown that patient characteristics on admission, like
co-morbidities, age and severity of illness do not fully explain the
increased long-term mortality. With lengthened stay in the ICU the
reason for stay in the ICU become increasingly less associated with
patient's original reason for admittance and increasingly associated
with persistent chronic illness with associated organ dysfunction [@23,
@24].

This being the case, this might be an explanation for the lack of
sigificance for ASA preinjury and ISS in the analysis. As stated above
severity of illness do not fully explain increased long-term mortality.
OFI may therefore not correlate fully with ASA preinjury and ISS
respectively.

However, a retrospective cohort has also shown that lower admission
illness severity and younger age were associated with earlier discharge
and therefore less associated with persistent critical illness, which
would then indicate a stronger association between ISS and ASA preinjury
to length of stay in the ICU as the patients who are in a better state
leave the hospital earlier and there is less possibilities for OFI. On
the other hand, the cohort also showed that lower comorbidity was
associated with persistent critical illness rather than early discharge
from the ICU --- supporting the opposite side of the argument [@25].

Several studies have also studied the association between different
factors which affect the length of stay in the ICU. Associations between
hypoglycaemia, use of any corticosteroids, use of any neuromuscular
agents, use of any benzodiazepines for more than 1 day, the need for
mechanical ventilation for \> 2 days, initiation of new dialysis and the
occurrence of new infection has been found with increased total and
5-year mortality [@23]. An association between bleeding in patients and
increased length of stay in the ICU [@27].

KLADD: APACHE is a physiological score and works best on critically ill
patients on a group level, but not as well on individual patients or on
relatively healthier patients [@28]...

@@

There are many studies on gender in association to health care. It has
previously been shown that there is no difference in willingness to
admit to the ICU between cases describing a man or woman in the
physician responses [@29]. However, studies also show that, among
severely injured patients, female gender is associated with lower
in-hospital mortality among those aged 16-44 years. Female gender is
also associated with decreased likelihood for ICU admission [@30]. Male
gender contributes more to poor outcomes in postoperative ICU patients,
and two-thirds of ICU resources are consumed by male patients [@31].
However, in our study we could not see any correlation between OFI and
one specific gender.

Preinjury ASA-PS score has been shown to be a reliable score for
predicting mortality after trauma, predicting readmission after
traumatic injury and for classifying comorbidity in trauma patients
[@32, @33, @34]. But why is ASA preinjury not associated with OFI? Could
it be because longer ICU stays are not associated with comorbidities to
the same extent? ...

The variable on duty was not significant in either analyses indicating
that the time of arrival of the patients did not affect their medical
treatment. Trying to reduce number of OFI can therefore not be solved by
increasing medical resources outside working hours. On-call sleeping has
been studied and it has been stated that sleeping between calls is
beneficial for waking performance. However, there might be a temporary
increase in performance impairment that should be taken into
consideration [@35]...

@@

ISS is an anatomical model, compared to RTS which is a physiological
model. Among trauma patients admitted to the ICU the physiological
models perform better than the anatomical models. However, patients with
low scores in the physiological model also require an anatomic analysis
of the injuries to determine their severity [@36]. NISS has been found
to be better than ISS in predicting the ICU admission of trauma
patients, as well as predicting prolonged ICU length of stay [@37]. This
could be an explanation to why ISS is not significant in the results of
this analysis, as RTS is a better model for patients admitted to the
ICU.

### Strengths and limitations

The strengths of this study was that it is a national registry-based
study which means that the quality of the data is reliable...

@@

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna as a base for choosing patient cases for the mortality
and morbidity review. Using audit filters to select patients for the
conference is associated with high false positive rates, from 24% to 80%
[@38]. Even though the use of audit filters is associated with high
false positive rates, there is still a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI that occurred later on during the
hospital stay.

@@

Another limitation is the combination of data from the emergency
department and prehospital values when the previous one's were missing.

@@

Examples of limitations:

-   Theoretical: limits the scope, depth, or applicability of a study.

-   Methodological: limits the quality, quantity, or diversity of the
    data.

-   Empirical: limits the representativeness, validity, or reliability
    of the data.

-   Analytical: limits the accuracy, completeness, or significance of
    the findings.

-   Ethical: limits the access, consent, or confidentiality of the data.
